Supplement to Global Distribution of Unintended Pregnancies and Contraceptive Usage Project

### 2024-04-19 code/eda-code/eda.qmd
Part 5 Updates
Contributed to strengths/limitations and added labels to the summary tables so there are no variable names.

# Create table 
predictortable <-selected_variables %>%
  tbl_summary(
    by = region,
    label = list(wra ~ "Number of women of reproductive age, 15-49"),
    type = all_continuous() ~ "continuous2",
    statistic = list(all_continuous() ~ c("{mean}","{min}, {max}"))) %>%
modify_header(label ~ "**Predictor Variable**") %>%
modify_caption("**Table 1. Predictor Variables**") %>%
modify_spanning_header(all_stat_cols() ~ "Region")

>>>>

# Modify region variable type for proper labeling 
selected_variables$region <- factor(selected_variables$region, labels = c("Africa", "Asia", "Europe", "Latin America/the Caribbean", "Oceania"))

# Create table 
predictortable <-selected_variables %>%
  tbl_summary(
    by = region,
    label = list(wra ~ "Number of women of reproductive age, 15-49",
                 starts_with("pct_upreg_sarc" ) ~ "Unintended Pregnancies due to SARC Usage (%)",
                 starts_with("pct_currentlymarried" ) ~ "Currently Married Women (%)",
                 starts_with("pct_nevermarried" ) ~ "Never Married Women (%)",
                 starts_with("rate_matdeaths") ~ "Rate of Maternal Deaths (per 100,000 live births)",
                 starts_with("rate_matdeaths_upreg") ~ "Rate of Maternal Deaths due to Unintended Pregnancies",
                 starts_with("rate_matdeaths_abs") ~ "Rate of Maternal Deaths due to Abortions"
                 ),
    type = all_continuous() ~ "continuous2",
    statistic = list(all_continuous() ~ c("{mean}","{min}, {max}"))) %>%
modify_header(label ~ "**Predictor Variable**") %>%
  modify_header(
    starts_with("region") ~ "**Region (1-Africa, 2-Asia, 3-Europe, 4-Latin America/the Caribbean, 5-Oceania)**"
  ) %>%
modify_caption("**Table 1. Summary Table**") %>%
modify_spanning_header(all_stat_cols() ~ "Region")

### 2024-04-19 products/manuscript/Manuscript.qmd
Part 5 Updates
Contributed to strengths/limitations and added labels to the summary tables so there are no variable names.

After deciding on a single outcome variable, we created a summary table with that outcome and the seven predictors we will use to predict it, which can be seen in @tbl-1. [insert explanation here]
```{r}
#| message: false
#| label: tbl-1
#| tbl-cap: "Summary Table Of Numeric Variables"
#| echo: FALSE
predictortable<- readRDS(here("results", "tables", "predictortable.rds"))
predictortable
```

@@

Lastly, we explored the predictor variables that will be used in our models, and the graphs are shown in @fig-5 - @fig-10. The Africa region accounts for most of the observations with about 53 countries, followed by Asia that includes about 35 countries. Latin America/the Caribbean accounts for about 25 countries. Europe and Oceania account for the least number of observations. The distribution of women currently married is approximately normal with an average around 57% and no major outliers. The distribution of women never married is slightly skewed to the right with an average around 30%, but there are a few outliers greater than 60%. The distribution of the total maternal mortality rate is highly skewed to the right with a couple outliers. We would have expected the highest rates to belong to India and China because they have the most people, but it is Chad, Sierra Leone, and South Sudan that account for the highest rates of total maternal mortality, which each belong to the African region. The distribution of the maternal mortality rate from unintended pregnancies is highly skewed to the right with a couple outliers, which is to be expected based on the trend shown in the previous histogram. Chad and South Sudan also account for the highest rates of maternal mortality in this category. The distribution of maternal mortality rate from abortions is also highly skewed to the right. The outliers are accounted for by Gambia and Mauritania in this predictor, which both belong to the African region.

@@

Since we determined that the full model with all 5 predictors performs best, we graphed the predicted versus observed values using this model with both the testing and training data to further examine model performance. As seen in @fig-11, the training model predictions, shown in teal, hover pretty closely around the expected relationship between the outcome and the 5 predictors, which is depicted as the 45 degree line. The testing model predictions, shown in coral, are much more extreme than the training model predictions because they are farther from the 45 degree line. While there is some overlap between the two sets of predictions, it is likely that our model is overfitting the data. 
```{r}
#| message: false
#| label: fig-11
#| fig-cap: "Evaluating the final model performance on test data reveals that the model is strongly overfitting the data. The testing model predictions are shown in coral, and the variation in these observations is much higher compared to the training model predictions, shown in blue. Ideally, there would be complete overlap between the testing and training data predictions following the black dotted line that represent the ideal relationship between observed and predicted values. The majority of testing data predictions do not overlap with the training data prediction in this case, and the testing data predictions are father away from the preferred relationship shown by the black dotted line."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "fullmodelpred.png"))
```
We went one step further to examine the uncertainty in the full model predictions using the bootstrap method with 100 resamples of the training data. The figure shown in @fig-12 displays the 95% confidence intervals, shown as dark blue lines, for the predictions made by the model. The confidence intervals are narrow, which shows there is minimal uncertainty in the model prediction. For the predictions that do not fall on the expected 45 degree line, the confidence intervals do not intersect with the line either. These findings increase the validity of the conclusion that the model is overfitting the data because there is not much uncertainty in the predictions. 
```{r}
#| message: false
#| label: fig-12
#| fig-cap: "Evaluation of uncertainty in the full model predictions reveals decent overlap between the point estimates, shown in black triangles, and the median prediction observation, shown in coral circles. The error bars, shown in blue, represent 95% confidence intervals of the median estimates, and most of the confidence intervals contain the point estimates. This provides confidence in the certainty of the model's predictions, but this is not very meaninful considering how the full model overfits the data."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "bootstrappred.png"))
```

Since we have evidence that our model overfits the data, the last technique we want to use to explore our full model is Lasso regression. This technique could reveal predictors that should not be included in our final model, which could reduce the risk of overfitting. As seen in @tbl-7, the LASSO regression coefficients are much smaller than the first full model we created without the regularization penalty, but all 5 predictors remain in the model. The `sarc_standard` predictor still has the strongest relationship with the outcome. The percent of women currently married, rate of maternal deaths due to unintended pregnancies, and rate of maternal deaths to due to abortion all have weak negative associations with the outcome. This process confirms that our final model should contain all 5 predictors. 
```{r}
#| message: false
#| label: tbl-7
#| fig-cap: "LASSO regression results confirm the findings of the multiple linear regression model, but each of the coefficients has been shrunk as a result of the penalty parameter of 0.1. The `sarc_standard` predictor is still the most strongly related to the outcome, followed by the European region. The LASSO penaltty decreased the effect size of the Oceania region even further, and the smallest coefficient is now assigned to the percent of currently married women, which is a shift from the multiple linear regression model. The rates of maternal mortality due to unintended pregnancy and abortion are very weak predictors in this model as well."
#| echo: FALSE
readRDS(here("results","tables", "resulttable16.rds")) 
```
It is important to tune the hyperparameters associated with LASSO regression to determine the model that fits best, which is shown in @fig-13. As the regularization penalty increases past 1, we see that the RMSE value increases exponentially. The lowest RMSE for the model is slightly greater than 5 when the regularization penalty is less than 1. When the regularization penalty increases to 10, the RMSE increases to almost 11. Since we originally chose 0.1 as our penalty parameter, we found one of the best performing models with the LASSO regression based on an RMSE value slightly greater than 5, which i comparable to the RMSE found in the first regression model with all 5 predictors. 
```{r}
#| message: false
#| label: fig-13
#| fig-cap: "Parameter tuning for the LASSO regression reveals that the parameter should be smaller than 1 to ensure the lowest RMSE performance metric value. The RMSE value increases exponentially when the penalty parameter exceeds 1, so this provides confidence in our choice of using 0.1 as the penalty parameter in the prevoiusly displayed LASSO regression model. There is no evidence to use a smaller penalty parameter than 0.1 because it does not result in a lower performance metric."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "lassoparameters.png"))
```

After creating the random forest model, we found that the RMSE and MAE values are exactly the same because the same 5 predictors are used for the outcome, but the R-squared value shows that the random forest model performs slightly better. Since a higher R-squared value represents smaller differences between the observed and predicted values, we can conclude that this model is predicting our outcome with slightly higher accuracy. The plot of the predicted versus observed values shown in @fig-14 displays stronger correlation for predicted values compared to the testing and training done with cross-validation previously.  
```{r}
#| message: false
#| label: fig-14
#| fig-cap: "The graph of predicted versus observed values in the random forest model shows less deviation from the expected relationship visualized by the black dotted line in comparison to the multiple linear regression model displayed previously. While there are still several outliers as the predicted values increase past a value of 20, the observations hover around the expected relationship with conistency for values less than 10. While it is still likely that the random forest model is overfitting because it uses the same predictors, it appears that this model has higher predictive accuracy."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "randomforestpred.png"))
```

Again, it is crucial to tune hyperparameters associate with machine learning models. Based on the red line displayed in @fig-15 that shows the lowest possible RMSE score, our random forest model performs the best with a node size of 1 and 4 predictors. This begs the question of which one of the 5 predictors included in our model should be dropped to improve model performance because all previous performance metrics suggested that including all 5 predictors created the best model. However, there is a minimal increase in RMSE when all five predictors are included in the model, but the results suggest that a full model should have a node size of 4 to achieve the lowest RMSE.
```{r}
#| message: false
#| label: fig-15
#| fig-cap: "After tuning two parameters for the random forest model, it is clear that the best model based on the lowest RMSE performance metric value has a minimal node size of 1 for 4 out of the 5 models. This graph suggests that the final model should contain 4 predictors to achieve the best performance metric, but it does not provide information on which parameter should be dropped from the model. Interestingly, the best model with all 5 predictors has a minimal node size of 4."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "randomforesttune.png"))
```

@@

## Strengths and Limitations
_Discuss what you perceive as strengths and limitations of your analysis._
This study relied on secondary data from the Guttmacher Institute's Adding It Up project. This means that we were limited by the data that was available and unable to collect our own data. Also, the study was cross-sectional, which means that it only looked at data from one point in time. This makes it difficult to establish causal relationships between the variables.

Despite these limitations, our study provides valuable insights into the factors influencing unintended pregnancy rates in low- and middle-income countries. The findings suggest that a multifaceted approach is needed to address unintended pregnancy. This approach should include increasing access to SARC and LARC methods, improving education and counseling on contraception use, and addressing regional disparities in access to healthcare services and family planning programs.

## Conclusions
_What are the main take-home messages?_
This study investigated the association between unintended pregnancy rates and short-acting reversible contraception (SARC) use across low- and middle-income countries. The findings suggest that SARC use, regional variations, and marital status are all important factors influencing unintended pregnancy rates.

@@

# References
_Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end_

>>>>

After deciding on a single outcome variable, we created a summary table with that outcome and the seven predictors we will use to predict it, which can be seen in @tbl-1. [insert explanation here]
```{r}
#| message: false
#| label: tbl-1
#| tbl-cap: "Summary Table Of Continuous Variables by Region"
#| echo: FALSE
predictortable<- readRDS(here("results", "tables", "predictortable.rds"))
predictortable
```

@@

Lastly, we explored the predictor variables that will be used in our models. The Africa region accounts for most of the observations with about 53 countries, followed by Asia that includes about 35 countries. Latin America/the Caribbean accounts for about 25 countries. Europe and Oceania account for the least number of observations. The distribution of women currently married is approximately normal with an average around 57% and no major outliers. The distribution of women never married is slightly skewed to the right with an average around 30%, but there are a few outliers greater than 60%. The distribution of the total maternal mortality rate is highly skewed to the right with a couple outliers. We would have expected the highest rates to belong to India and China because they have the most people, but it is Chad, Sierra Leone, and South Sudan that account for the highest rates of total maternal mortality, which each belong to the African region. The distribution of the maternal mortality rate from unintended pregnancies is highly skewed to the right with a couple outliers, which is to be expected based on the trend shown in the previous histogram. Chad and South Sudan also account for the highest rates of maternal mortality in this category. The distribution of maternal mortality rate from abortions is also highly skewed to the right. The outliers are accounted for by Gambia and Mauritania in this predictor, which both belong to the African region.

@@

Since we determined that the full model with all 5 predictors performs best, we graphed the predicted versus observed values using this model with both the testing and training data to further examine model performance. As seen in @fig-6, the training model predictions, shown in teal, hover pretty closely around the expected relationship between the outcome and the 5 predictors, which is depicted as the 45 degree line. The testing model predictions, shown in coral, are much more extreme than the training model predictions because they are farther from the 45 degree line. While there is some overlap between the two sets of predictions, it is likely that our model is overfitting the data. 
```{r}
#| message: false
#| label: fig-6
#| fig-cap: "Evaluating the final model performance on test data reveals that the model is strongly overfitting the data. The testing model predictions are shown in coral, and the variation in these observations is much higher compared to the training model predictions, shown in blue. Ideally, there would be complete overlap between the testing and training data predictions following the black dotted line that represent the ideal relationship between observed and predicted values. The majority of testing data predictions do not overlap with the training data prediction in this case, and the testing data predictions are father away from the preferred relationship shown by the black dotted line."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "fullmodelpred.png"))
```
We went one step further to examine the uncertainty in the full model predictions using the bootstrap method with 100 resamples of the training data. The figure shown in @fig-7 displays the 95% confidence intervals, shown as dark blue lines, for the predictions made by the model. The confidence intervals are narrow, which shows there is minimal uncertainty in the model prediction. For the predictions that do not fall on the expected 45 degree line, the confidence intervals do not intersect with the line either. These findings increase the validity of the conclusion that the model is overfitting the data because there is not much uncertainty in the predictions. 
```{r}
#| message: false
#| label: fig-7
#| fig-cap: "Evaluation of uncertainty in the full model predictions reveals decent overlap between the point estimates, shown in black triangles, and the median prediction observation, shown in coral circles. The error bars, shown in blue, represent 95% confidence intervals of the median estimates, and most of the confidence intervals contain the point estimates. This provides confidence in the certainty of the model's predictions, but this is not very meaninful considering how the full model overfits the data."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "bootstrappred.png"))
```

Since we have evidence that our model overfits the data, the last technique we want to use to explore our full model is Lasso regression. This technique could reveal predictors that should not be included in our final model, which could reduce the risk of overfitting. As seen in @tbl-7, the LASSO regression coefficients are much smaller than the first full model we created without the regularization penalty, but all 5 predictors remain in the model. The `sarc_standard` predictor still has the strongest relationship with the outcome. The percent of women currently married, rate of maternal deaths due to unintended pregnancies, and rate of maternal deaths to due to abortion all have weak negative associations with the outcome. This process confirms that our final model should contain all 5 predictors. 
```{r}
#| message: false
#| label: tbl-7
#| fig-cap: "LASSO regression results confirm the findings of the multiple linear regression model, but each of the coefficients has been shrunk as a result of the penalty parameter of 0.1. The `sarc_standard` predictor is still the most strongly related to the outcome, followed by the European region. The LASSO penaltty decreased the effect size of the Oceania region even further, and the smallest coefficient is now assigned to the percent of currently married women, which is a shift from the multiple linear regression model. The rates of maternal mortality due to unintended pregnancy and abortion are very weak predictors in this model as well."
#| echo: FALSE
readRDS(here("results","tables", "resulttable16.rds")) 
```
It is important to tune the hyperparameters associated with LASSO regression to determine the model that fits best, which is shown in @fig-8. As the regularization penalty increases past 1, we see that the RMSE value increases exponentially. The lowest RMSE for the model is slightly greater than 5 when the regularization penalty is less than 1. When the regularization penalty increases to 10, the RMSE increases to almost 11. Since we originally chose 0.1 as our penalty parameter, we found one of the best performing models with the LASSO regression based on an RMSE value slightly greater than 5, which i comparable to the RMSE found in the first regression model with all 5 predictors. 
```{r}
#| message: false
#| label: fig-8
#| fig-cap: "Parameter tuning for the LASSO regression reveals that the parameter should be smaller than 1 to ensure the lowest RMSE performance metric value. The RMSE value increases exponentially when the penalty parameter exceeds 1, so this provides confidence in our choice of using 0.1 as the penalty parameter in the prevoiusly displayed LASSO regression model. There is no evidence to use a smaller penalty parameter than 0.1 because it does not result in a lower performance metric."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "lassoparameters.png"))
```

After creating the random forest model, we found that the RMSE and MAE values are exactly the same because the same 5 predictors are used for the outcome, but the R-squared value shows that the random forest model performs slightly better. Since a higher R-squared value represents smaller differences between the observed and predicted values, we can conclude that this model is predicting our outcome with slightly higher accuracy. The plot of the predicted versus observed values shown in @fig-9 displays stronger correlation for predicted values compared to the testing and training done with cross-validation previously.  
```{r}
#| message: false
#| label: fig-9
#| fig-cap: "The graph of predicted versus observed values in the random forest model shows less deviation from the expected relationship visualized by the black dotted line in comparison to the multiple linear regression model displayed previously. While there are still several outliers as the predicted values increase past a value of 20, the observations hover around the expected relationship with conistency for values less than 10. While it is still likely that the random forest model is overfitting because it uses the same predictors, it appears that this model has higher predictive accuracy."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "randomforestpred.png"))
```

Again, it is crucial to tune hyperparameters associate with machine learning models. Based on the red line displayed in @fig-10 that shows the lowest possible RMSE score, our random forest model performs the best with a node size of 1 and 4 predictors. This begs the question of which one of the 5 predictors included in our model should be dropped to improve model performance because all previous performance metrics suggested that including all 5 predictors created the best model. However, there is a minimal increase in RMSE when all five predictors are included in the model, but the results suggest that a full model should have a node size of 4 to achieve the lowest RMSE.
```{r}
#| message: false
#| label: fig-10
#| fig-cap: "After tuning two parameters for the random forest model, it is clear that the best model based on the lowest RMSE performance metric value has a minimal node size of 1 for 4 out of the 5 models. This graph suggests that the final model should contain 4 predictors to achieve the best performance metric, but it does not provide information on which parameter should be dropped from the model. Interestingly, the best model with all 5 predictors has a minimal node size of 4."
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "final-figures", "randomforesttune.png"))
```

@@

## Strengths and Limitations
There are several limitations for this analysis that must be addressed. First, this study relied on secondary data from the Guttmacher Institute's Adding It Up project. We were limited by the data that was available to us because we were unable to collect our own data. The study was cross-sectional, which means that it only looked at data from one point in time. This makes it difficult to establish causal relationships between the variables based on lack of information about temporality. The largest limitation of the study is the evidence of overfitting for our final model. While we had solid evidence from performance metrics for choosing this model, it does a fairly poor job of representing the data based on the cross-validated testing and training data predictions. Our recommendation is to explore more variables from the Adding It Up project to determine if there are better predictors for this outcome. Since we chose our predictors based on the literature, it is unlikely that there are other predictors that will more accurately predict the percentage of unintended pregnancies due to SARCs. Further research into the other contraceptive methods studied in the Adding It Up project, such as long acting reversible contraception, sterilization, and traditional methods, could provide clarity on the patterns that we discovered in this analysis. Due to limited time and experience, we were not able to examine all 4 of these outcomes in this project.

Despite these limitations, our study provides valuable insights into the factors influencing unintended pregnancy rates in low- and middle-income countries. We utilize cutting edge machine learning models to test 10 different model options, cross-validate 3 of the best performing models, and tune parameters for LASSO regression and random forest modeling of our final model. These robust techniques allowed for a thorough investigation of predictors and outcomes. Our findings suggest that a multifaceted approach is needed to address unintended pregnancy. This approach should include increasing access to SARC methods, improving education and counseling on contraception use, and addressing regional disparities in access to healthcare services and family planning programs.

## Conclusions
This study investigated the association between unintended pregnancy rates and short-acting reversible contraception (SARC) use across low- and middle-income countries. The findings suggest that SARC use, regional variations, and marital status are all important factors influencing unintended pregnancy rates.

@@

# References

### 2024-04-19 products/manuscript/supplement/Supplementary-Material.qmd
Part 5 Updates
Contributed to strengths/limitations and added labels to the summary tables so there are no variable names.

Any additional discussion regarding the supplementary material/findings.

# References

>>>>

The most helpful part of this supplementary material is the information about the other potential outcomes that can be studied including the percentage of unintended pregnancies due to long-acting reversible methods, sterilization, traditional methods, and no birth control method. The initial exploration into these variables will provide future researchers with an initial idea of the distribution of these variables. The additional exploratory analysis included above provides insight on how we standardized our population predictor and dealt with missing data, which increases the validity of our study methods. The details about the methods we chose add confidence that the analysis performed in the statistical-analysis.qmd was the correct approach for our data. Choosing robust techniques like the LASSO regression and bootstrapping add evidence to our thorough investigation of the outcome and 7 predictors.
